---
title: "Explore"
output: github_document
---

```{r}
library(lubridate)
library(tidyverse)
```


```{r}
bird <- readRDS("cache/bird.rds")
brd_df <- bird$brd_countdata
```



How do we use this sampling to construct a measure of species abundance over time?

Sampling occurs at discrete sites in discrete 6 minute events in which one or more observers count every bird they can observe in that interval.  Observers note the species ID, approximate distance to the bird (or cluster), number of birds in a cluster, method of identification (visual, singing, etc) and sometimes additional information (sex) during each minute-length interval of the 6 minute event.

I assume individual observers do not enter the same bird twice under two seperate identification methods, or that multiple observers do not both enter an observation for the same bird, but this can probably be confirmed in the protocol. 

A sampling event is defined then by a unique ID which indicates a point in space and time.  The  location  is given by a combination of pointID, which is unique only within a plot, and the  plotID, which includes the site.  Time is given as a startDate which may be shared across several events.  Thus for a given species, we can sum clusterSizes for each unique event to find the total count of species that was observed at that point in a given 6-minute interval.


```{r}
  ## sum bird (clusters) counted at each of the six minutes of the sampling event 
  ##(i.e. regardless of observerDistance and detectionMethod, or pointCountMinute)
brd_event <- brd_df %>% 
  filter(taxonRank == "species") %>%
  group_by(scientificName, pointID, siteID, plotID, eventID, startDate) %>% 
  summarize(counts_per_6min = sum(clusterSize)) %>% ungroup() 
  
## A sighting event is a 6min interval
## an eventID is plotID+pointID+startDate
## A plotID includes the siteID
```

How do we combine this over points, plots, etc for a density?  We could view the multiple events that occur at a given point as independent estimates of the density at that point and average them.  We could likewise consider the multiple points in a plot as independent estimates of the plot and average those, etc.

But we should probably be summing instead of averaging?  e.g. count up all the occurances in an area and divide by the 'size' of the area.  Though it is not clear that we can say if a given area (e.g. plot) was sampled completely -- plots could differ greatly in size.  


```{r}
## We can have many different points in the same plot
#brd_event %>% count(scientificName, siteID, plotID, startDate, sort=T)

## So we should average across those.
brd_plot <- brd_event %>%  
  group_by(scientificName, siteID, plotID, startDate) %>% 
  summarise(mean_count_per_6min = mean(counts_per_6min, na.rm = FALSE))

## Or at larger spatial aggregation: all plots in the site
brd_site <- brd_event %>%  
  group_by(scientificName, siteID, startDate) %>% 
  summarise(mean_count_per_6min = mean(counts_per_6min, na.rm = FALSE))

## Or across all of space: 
brd_total  <- brd_event %>%  
  group_by(scientificName, startDate) %>% 
  summarise(mean_count_per_6min = mean(counts_per_6min, na.rm = FALSE))

```

Most sampling takes place in a condensed season of several sampling days/weeks for just a few months of the year.  We thus should probably think of these as year averages.  

```{r}
brd_event %>% mutate(year = year(startDate), month = month(startDate)) %>% count(siteID, plotID, scientificName, year, month, sort = TRUE) 
```


```{r}

common_sp <- brd_total %>%
  group_by(scientificName) %>% 
  summarise(sightings = mean(mean_count_per_6min, na.rm = FALSE)) %>% 
  arrange(desc(sightings)) 

brd_total %>% 
  inner_join(common_sp%>% head(10)) %>% 
  ggplot(aes(startDate, mean_count_per_6min)) +
  geom_point() +
  facet_wrap(~scientificName, scales="free_y")

```


```{r}
brd_annual_site <- brd_site %>% mutate(year = year(startDate)) %>% ungroup() %>% 
  group_by(scientificName, year) %>% summarise(mean_count = mean(mean_count_per_6min, na.rm = FALSE)) 


brd_annual_totals <- 
brd_total %>% mutate(year = year(startDate)) %>% ungroup() %>% 
  group_by(scientificName, year) %>% summarise(mean_count = mean(mean_count_per_6min, na.rm = FALSE)) 
```






Some species have crazy high counts per 6 min, perhaps not believable: 

```{r}
brd_annual_totals %>% inner_join(head(common_sp,6)) %>% 
    ggplot(aes(year, mean_count)) + geom_point() + geom_line() + facet_wrap(~scientificName, scales="free_y")
```



Focus on timeseries: 
```{r}
most_data <- brd_annual_totals %>%
    inner_join(common_sp) %>% count(scientificName, sort = TRUE) %>% head(6) %>% select(scientificName)
```

```{r}
brd_annual_totals %>% inner_join(most_data) %>% 
    ggplot(aes(year, mean_count)) + geom_point() + geom_line() + facet_wrap(~scientificName, scales="free_y")
```



