---
output:
  github_document:
    df_print: tibble

---

The NEON bird sampling protocol is clearly designed with modern distance-sampling
based protocols in mind.  



```{r message=FALSE}
library(lubridate)
library(tidyverse)
library(Distance)
library(knitr)

## remotes::install_github("cboettig/neonstore")

# use cache since model fitting can take a while
opts_chunk$set(cache=TRUE)
```

```{r}
## ick, manually force columns bc vroom is confused about NEON's
## non-standard timestamp notation 
s <- cols(
  uid = col_character(),
  namedLocation = col_character(),
  domainID = col_character(),
  siteID = col_character(),
  plotID = col_character(),
  plotType = col_character(),
  pointID = col_character(),
  startDate = col_character(),
  eventID = col_character(),
  pointCountMinute = col_double(),
  targetTaxaPresent = col_character(),
  taxonID = col_character(),
  scientificName = col_character(),
  taxonRank = col_character(),
  vernacularName = col_character(),
  family = col_character(),
  nativeStatusCode = col_character(),
  observerDistance = col_double(),
  detectionMethod = col_character(),
  visualConfirmation = col_character(),
  sexOrAge = col_character(),
  clusterSize = col_double(),
  clusterCode = col_character(),
  identifiedBy = col_character()
)

brd_countdata <- neonstore::neon_read("brd_countdata", col_types = s)
```

Let's re-arrange the data slightly so that is more suitable for input in the widely used `Distance` package.  

First, let's make all `pointID`s identify unique points in space
by including the ID of the plot (which includes the siteID already) 
in the `pointID`  

If we want annual estimates of the abundance of each species, maybe
we should round the `startDate` to year, allowing us to treat the all visits to the same point in the same year as the effort that year.

Lastly, we'll ignore the issue of Area for the moment and consider density estimates alone, which should be reasonable for forecasting (after all the Area, whatever it is, will remain fixed.)

```{r}
bird_ds <- brd_countdata %>% 
  mutate(pointID = paste(plotID, pointID, sep="-")) %>%
  mutate(year = lubridate::year(startDate), Area = 1)
```

Now rename columns to match `ds` columns, or reflect grouping variables and interesting covariates.

```{r}
ds_data <- 
  bird_ds  %>% 
  mutate(Region.Label = paste(siteID, year, sep="-")) %>%
  select(scientificName, 
         year,
         siteID,
         plotID,
         Region.Label,
         obs_id = identifiedBy,
         detection_method = detectionMethod,
         distance = observerDistance, 
         Sample.Label = eventID,
         size = clusterSize) %>%
  mutate(obs_id = as.factor(obs_id),
         Effort = 1,
         detection_method = as.factor(detection_method))
```

`eventID` gives identifies visit to a point sampler at a given time, this seems like a reasonable `Sample.Label` (if one wanted to ignore temporal differences `plotID` would be acceptable to identify only the point sampler location).

`Region.Label` will give the level at which we wish to estimate abundance. A reasonable target for this might be a Neon site in a given year, this ensures replication at each location (since there are multiple points) so variance estimation will likely be better behaved.

`Effort` is number of visits to sample unit (`Sample.Label`), so is 1 for all entries.

 
## Selecting focal species

```{r}
brd_countdata %>% select(siteID, scientificName) %>%
  distinct() %>% count(scientificName, sort = TRUE) %>% head()


brd_countdata %>% select(plotID, scientificName) %>%
  distinct() %>% count(scientificName, sort = TRUE) %>% head()

```

Let's mourning doves as they are common across sites and plots.

```{r}
doves <- ds_data %>% 
  filter(scientificName == "Zenaida macroura") %>% 
  as.data.frame()
```

It seems reasonable to pool data over all years and expect detection to be similar (we should check this at some point by adding a covariate for year!)

First plot the histogram of distances:

```{r}
hist(doves$distance, xlab="Distance (m)", breaks=20)
```

Fit a simple model, truncating at 500m to begin with:

```{r}
df_dove_hn_cos_500 <- Distance::ds(doves, transect = "point", truncation=500)
```

Plot that (both the detection function and PDF of distances):
```{r}
par(mfrow=c(1,2))
plot(df_dove_hn_cos_500)
plot(df_dove_hn_cos_500, pdf=TRUE)
```

PDF makes it look like there is some overfitting here -- those extra adjustment terms are probably overfitting the tail.

Rough rule of thumb that we want probability of detection to be $\approx 0.15$ at the truncation distance, so we can truncate a lot more here:

```{r}
df_dove_hn_cos_150 <- Distance::ds(doves, transect = "point", truncation=150)
```

Plot that:
```{r}
par(mfrow=c(1,2))
plot(df_dove_hn_cos_150)
plot(df_dove_hn_cos_150, pdf=TRUE)
```

Maybe an issue here that the 2nd bin in this histograms looks too tall. This may be because observers are flushing birds? One way around this is to bin the distances.

Then try to fit the model again using the `cutpoints=` argument (expanding the truncation a wee bit):

```{r}
df_dove_hn_cos_200_bin <- Distance::ds(doves, transect = "point", truncation=200,
                                       cutpoints=seq(0, 200, by=20))
```

```{r}
par(mfrow=c(1,2))
plot(df_dove_hn_cos_200_bin)
plot(df_dove_hn_cos_200_bin, pdf=TRUE)
```

That looks better behaved.

Could also try with hazard-rate detection function:

```{r}
df_dove_hr_cos_200_bin <- Distance::ds(doves, transect = "point", truncation=200,
                                       key="hr", cutpoints=seq(0, 200, by=20))
```

```{r}
par(mfrow=c(1,2))
plot(df_dove_hr_cos_200_bin)
plot(df_dove_hr_cos_200_bin, pdf=TRUE)
```

Compare via AIC?

```{r}
AIC(df_dove_hn_cos_200_bin, df_dove_hr_cos_200_bin)
```

Not much in it! What about goodness of fit?

```{r}
ds.gof(df_dove_hn_cos_200_bin)
ds.gof(df_dove_hr_cos_200_bin)
```


# Covariates

What if group size affects detectability?

```{r}
df_dove_hn_size_200_bin <- Distance::ds(doves, transect = "point", truncation=200,
                                        formula=~size, cutpoints=seq(0, 200, by=20))

df_dove_hr_size_200_bin <- Distance::ds(doves, transect = "point", truncation=200,
                                        key="hr", formula=~size,
                                        cutpoints=seq(0, 200, by=20))
```

(We *could* think about other covariates here. `obs_id` would be an option but it has `r length(unique(doves$obs_id))` levels, that model takes a very long time to fit.)

Which of those looks best via AIC?

```{r}
AIC(df_dove_hn_cos_200_bin, df_dove_hr_cos_200_bin, df_dove_hr_size_200_bin, df_dove_hn_size_200_bin)
```

Plot, 

```{r}
par(mfrow=c(2,2))
plot(df_dove_hr_size_200_bin, main="hr_size", pdf=TRUE)
plot(df_dove_hn_size_200_bin, main="hn_size", pdf=TRUE)
plot(df_dove_hr_cos_200_bin, main="hr", pdf=TRUE)
plot(df_dove_hn_cos_200_bin, main="hn", pdf=TRUE)
```


# Results tables

We can compare multiple models:
```{r}
res_tab <- summarize_ds_models(df_dove_hn_cos_200_bin, df_dove_hr_cos_200_bin,
                               df_dove_hr_size_200_bin, df_dove_hn_size_200_bin, output="plain")
```

And `kable` them:

```{r results="asis"}
kable(res_tab)
```


# Per year abundance estimates 

We can do this in various ways (including via setting the `Region.Label` appropriately above and estimate at the same time as fitting). `Distance` also provides `dht2` which allows for more complex abundance estimation. Let's use `df_dove_hr_cos_200_bin` as our detection function for these estimates.


```{r}
# set the stratum areas to just be the covered area for that year
doves_areas <- doves %>%
  select(Region.Label, Sample.Label, Effort) %>%
  distinct() %>%
  mutate(Area=sum(Effort) * df_dove_hn_cos_200_bin$ddf$meta.data$width^2 * pi)
# only 1 value here so we can cheat
doves$Area <- doves_areas$Area[1]


year_site_abund <- dht2(df_dove_hn_cos_200_bin, flatfile=doves,
                        strat_formula=~Region.Label)

```

Plot some trends?

```{r}
library(ggplot2)
year_site_abund_plot <- as.data.frame(attr(year_site_abund, "density"))
year_site_abund_plot$year <- as.numeric(sub("^.{4}-", "", year_site_abund_plot$Region.Label))
year_site_abund_plot$site <- sub("-\\d{4}", "", year_site_abund_plot$Region.Label)

ggplot(year_site_abund_plot) +
  geom_line(aes(x=year, y=Density, colour=site, group=site)) +
  theme_minimal()
```

