---
output: 
  github_document:
    df_print: tibble
---

The NEON bird sampling protocol is clearly designed with modern distance-sampling
based protocols in mind.  



```{r message=FALSE}
library(lubridate)
library(tidyverse)
library(Distance)

## remotes::install_github("cboettig/neonstore")
```

```{r}
## ick, manually force columns bc vroom is confused about NEON's
## non-standard timestamp notation 
s <- cols(
  uid = col_character(),
  namedLocation = col_character(),
  domainID = col_character(),
  siteID = col_character(),
  plotID = col_character(),
  plotType = col_character(),
  pointID = col_character(),
  startDate = col_character(),
  eventID = col_character(),
  pointCountMinute = col_double(),
  targetTaxaPresent = col_character(),
  taxonID = col_character(),
  scientificName = col_character(),
  taxonRank = col_character(),
  vernacularName = col_character(),
  family = col_character(),
  nativeStatusCode = col_character(),
  observerDistance = col_double(),
  detectionMethod = col_character(),
  visualConfirmation = col_character(),
  sexOrAge = col_character(),
  clusterSize = col_double(),
  clusterCode = col_character(),
  identifiedBy = col_character()
)

brd_countdata <- neonstore::neon_read("brd_countdata", col_types = s)
```



Let's re-arrange the data slightly so that is more suitable for input in the widely used `Distance` package.  


First, let's make all `pointID`s identify unique points in space
by including the ID of the plot (which includes the siteID already) 
in the `pointID`  

If we want annual estimates of the abundance of each species, maybe
we should round the `startDate` to year, allowing us to treat the all visits to the same point in the same year as the effort that year.

Lastly, we'll ignore the issue of Area for the moment and consider density estimates alone, which should be reasonable for forecasting (after all the Area, whatever it is, will remain fixed.)

```{r}
bird_ds <- brd_countdata %>% 
  mutate(pointID = paste(plotID, pointID, sep="-")) %>%
  mutate(year = lubridate::year(startDate), Area = 1)
```


Not really sure on this, but I'm going to set sampling effort as 
the number of visits to a given point?

```{r}
effort <- bird_ds %>% 
  select(pointID,year) %>% 
  distinct() %>% 
  count(pointID)
```

Join the effort data back on by pointID

```{r}
ds_data <- bird_ds %>%
  left_join(rename(effort, Effort = n), by = "pointID")
```

Now rename columns to match `ds` columns, or reflect grouping variables (species, year).

```{r}
ds_data <- 
  ds_data  %>% 
  select(scientificName, 
         year,
         distance = observerDistance, 
         Sample.Label = eventID, 
         Region.Label = plotID, 
         size = clusterSize)
```


 
## Selecting focal species

```{r}
brd_countdata %>% select(siteID, scientificName) %>%
  distinct() %>% count(scientificName, sort = TRUE) %>% head()


brd_countdata %>% select(plotID, scientificName) %>%
  distinct() %>% count(scientificName, sort = TRUE) %>% head()

```

Cool, morning doves are common across sites and plots, let's try something with those.

```{r}
doves <- ds_data %>% 
  filter(scientificName == "Zenaida macroura", year == "2015") %>% 
  as.data.frame()
```




```{r}
abund <- Distance::ds(doves, transect = "point")
```










Let's hack the `Distance::ds` print method to provide a `data.frame` return object instead...


```{r}

## Tidy the `ds` return object
 estimate_table <- function(abund) {
  abund_summary <- summary(abund)
  x <- abund_summary$ds
  if(!is.null(x$Nhat)){
    parameters = data.frame(Estimate = c(x$average.p,x$Nhat))
    row.names(parameters) = c("Average p", "N in covered region")
    if(!is.null(x$average.p.se)){
      parameters$SE = c(x$average.p.se,x$Nhat.se)
      parameters$CV = parameters$SE/parameters$Estimate
    }
  } else {
    parameters = data.frame(Estimate = c(x$average.p))
    row.names(parameters) = c("Average p")
    if(!is.null(x$average.p.se)){
      parameters$SE = c(x$average.p.se)
      parameters$CV = parameters$SE/parameters$Estimate
    }
  }
  out <- tibble::rownames_to_column(parameters)
out
}
```

We can not get a table summary of the results:

```{r}
estimate_table(abund)
```

Now we can do multiple years at once: 

```{r}

estimate_ds <- function(df, ...){
    m <- ds(as.data.frame(df), transect = "point")
    estimate_table(m)
}

doves_abund <- ds_data %>% 
  filter(scientificName == "Zenaida macroura") %>% 
  group_by(year) %>%
dplyr::group_modify(estimate_ds, keep = TRUE)

doves_abund
```

