---
title: "Tick Abundance"
output: 
  github_document:
    df_print: tibble
---

```{r setup,message=FALSE}
library(lubridate)
library(tidyverse)
```


```{r}
# neonstore::neon_download("DP1.10093.001")
tck <- neonstore::neon_read("tck_taxonomyRaw-expanded")
sites <- neonstore:::neon_sites() 

```


```{r}
tck_ts <- tck %>% 
  mutate(siteID = gsub("(\\w+{4})_\\d{3}", "\\1", plotID)) %>%  # NEON forgot this field this time...
  select(collectDate, siteID, scientificName, individualCount) %>%
  mutate(month = format(collectDate, "%Y-%m")) %>% 
  mutate(month = as.Date(paste(month, "01", sep="-"))) %>%
  group_by(scientificName, month, siteID) %>%
  summarize(count = sum(individualCount, na.rm = TRUE)) %>%
  ungroup()
```

## Counts

First let's take a quick look at the raw counts data.  This is less meaningful than abundance,
since we would need to account for detection probability, and effort is not entirely constant over
time or necessarily equal across sites.  Details of the estimation of beetle density will need to
take into account the specifics of the pitfall sampling design.  For now, let's consider only the
raw counts, which are much simpler to work with and free from assumptions required to estimate abundance:


To start, here is cumulative counts across all species: shows an increase which is no doubt connected 
to increased sampling effort as sites come online, along with an obvious seasonal 
pattern...


```{r}
totals <- tck_ts %>% 
  group_by(month) %>%
  summarize(count = sum(count, na.rm = TRUE)) %>%
  ungroup()

totals %>% ggplot(aes(month, count)) + geom_line() + geom_point()
```

2018 doesn't have a huge spike in number of site-months being reported though...

```{r}
tck_ts %>% count(month, sort = TRUE)
```






